Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-20 analysis/paper/paper.Rmd
Do a CA using only the flint data

# Find date to hold the central tendency date
centdate <- (sum(range(lithics$mean_date)) / 2) / 1000
```
# Results

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves. The non-flint material on the earliest sites appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. These findings are in large part in line with previous research [e.g. @bang-andersen2003], although highlight the importance of non-flint materials for the production of projectiles.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these later sites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370, @glørstad2011].

@@

In addition, Nedre Hobekk 2 represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite. The use of metarhyolite is typically seen as a feature of the end of the EM and the MM, but is evidently not as prominent a part of other sites that are contemporaneous with Nedre Hobekk 2.

@@

```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity."}
# , fig.height = 7, fig.width = 10

@@

Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r = *`r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. The correlation is stronger between the mean site age and RFSL (*r = *`r round(secfreqcor, 2)`), than between mean site age and VDL (*r = *`r round(denscor, 2)`).  As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data. While the negative correlation is slightly less marked when only the flint data is considered (*r = *`r round(wabicorflint, 2)`), the general pattern is the same. The relationship between mean site age and RFSF is even stronger (*r = *`r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the VDF is not temporally contingent (*r = *`r round(denscorflint, 2)`). This could perhaps indicate that when non-flint materials make up a higher share of the assemblages, as is especially the case for some of the later Mesolithic sites, expediency might to a larger extent be associated with the non-flint component of the assemblages rather than the flint material. This point is also returned to below where the temporal dimension of the relationship is explored further.

@@

To get at this temporal trend a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot). Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This stabilises around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic. The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.

@@

While the relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here arrived at by means of an exploratory approach, some inferential merit can be achieved by invoking what has been termed consilience. Consilience involves "the interlocking or coherence of causal explanations across multiple problem domains" through a clear operationalisation of explanatory terms and concepts [@clark2009, 30]. Thus, while by no means a firm , the clear overlap in results presented here and those repeatedly reported by does to the general applicability of the measures, and offers an initial indication that they are capturing the social dimensions of interest also in a Norwegian Mesolithic setting. While it is not within the scope of this paper to consider these comparisons to any wider extent, the VDL for all lithics ranges from `r round(max(lithics2$lithic_dens))` lithics per m^3, down to `r round(min(lithics2$lithic_dens))` which appears to be relatively high when compared to Late Pleistocene sites from western Eurasia. Similarly, the frequency of retouched artefacts ranges from `r round(min(lithics2$secondary_freq)*100, 1)`--`r round(max(lithics2$secondary_freq)*100, 1)` % which is relatively low when compared to the same data. This could, in sum, indicate that the coastal Mesolithic sites of South-Eastern Norway have generally been associated with a low degree of mobility when compared to these context, a proposition that would be in line with both ethnographic and archaeological research that highlights the possible role marine resources can play.

The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises from around 7000 BCE for the rest of the Mesolithic. This pattern is evident in both the flint data and when all lithics are treated in aggregate. The variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. Referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages. As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages. The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps. The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel out. Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern, while the high expediency on some sites when considering all lithics in aggregate in the later parts of the Mesolithic does speak to variability in site use. The nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential-logistic continuum. A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014]. However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.

>>>>

# Find date to hold the central tendency date
centdate <- (sum(range(lithics$mean_date)) / 2) / 1000
```

# Results

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to larger extent are characerised by core fragments, both in flint and non-flint materials, than cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centered around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. The single red point in the upper right quadrant of the row plot is Nedre Hobekk 2, which represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite. The use of metarhyolite and smoky quartz for the production of axes is present at other sites as well, but is evidently not as prominent a part of these assemblages. It should also be noted that the later Mesolithic sites that are also suppressing this pattern somewhat. In sum, the findings are in large part in line with previous research [e.g. @bang-andersen2003], although perhaps highlighting the relative importance of non-flint materials in south-eastern which, while also present, is typically not as marked in other parts of southern Norway [@breivik2018, 184]

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages and non-flint projectiles. The outer end of the first dimensions is dominated by later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among these latsites. Although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot. This could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous research [e.g. @eigeland2015, 370, @glørstad2011].

@@

@@

```{r pca-analysis, fig.cap = "PCA using variables related. Note that details on the transformation of the variables has been left out of the plot for clarity.", fig.height = 8}
# fig.width = 10

@@

Figure \@ref(fig:wabi)A illustrates the negative correlation between the two variables (*r =* `r round(wabicor, 2)`) while also displaying a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. The correlation is stronger between the mean site age and RFSL (*r =* `r round(secfreqcor, 2)`), than between mean site age and VDL (*r =* `r round(denscor, 2)`). As variable non-flint availability and workability has also been suggested to potentially impact these measures [@manninen2014], Figure \@ref(fig:wabi)B displays the same relationship, but exclusively for the flint data. While the negative correlation is slightly less marked when only the flint data is considered (*r =* `r round(wabicorflint, 2)`), the general pattern is the same. The relationship between mean site age and RFSF is even stronger (*r =* `r round(secfreqcorflint, 2)`), but as indicated by the more spread out distribution along the y-axis, the VDF is not temporally contingent (*r =* `r round(denscorflint, 2)`). As was also indicated by the CA, this follows from the fact that non-flint materials make up a higher share of the assemblages for some of the later Mesolithic sites, and is a point returned to below where the temporal dimension of the relationship between VDL and RFSL is explored further.

@@

To get at this temporal trend, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites as provided in Figure \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in Figure \@ref(fig:simplot). Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This stabilises around 7000 BCE and remains stable without any major fluctuations for the rest of the Mesolithic. The variation in degree of curation is also markedly higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.

@@

While the relevance of the relationship between frequency of secondarily worked lithics and density of lithics was here arrived at by means of an exploratory approach, some inferential merit can be achieved by invoking what has been termed consilience. Consilience involves "the interlocking or coherence of causal explanations across multiple problem domains" through a clear operationalisation of explanatory terms and concepts [@clark2009, 30]. Thus, while by no means a firm , the clear overlap in results presented here and those repeatedly reported by does to the general applicability of the measures, and offers an initial indication that they are capturing the social dimensions of interest also in a Norwegian Mesolithic setting. While it is not within the scope of this paper to consider these comparisons to any wider extent, the VDL for all lithics ranges from `r round(max(lithics2$lithic_dens))` lithics per m\^3, down to `r round(min(lithics2$lithic_dens))` which appears to be relatively high when compared to Late Pleistocene sites from western Eurasia. Similarly, the frequency of retouched artefacts ranges from `r round(min(lithics2$secondary_freq)*100, 1)`--`r round(max(lithics2$secondary_freq)*100, 1)` % which is relatively low when compared to the same data. This could, in sum, indicate that the coastal Mesolithic sites of South-Eastern Norway have generally been associated with a low degree of mobility when compared to these context, a proposition that would be in line with both ethnographic and archaeological research that highlights the possible role marine resources can play.

The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises from around 7000 BCE for the rest of the Mesolithic. This pattern is evident in both the flint data and when all lithics are treated in aggregate. The variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. Referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages. As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages. Thus, the degree of expediency for assemblages dominated by non-flint materials might be somewhat exaggerated when the non-flint material is included, while its exlusion seems to underplay. As the measure and general tendencies appear relatively robust to this effect, this it not explored further here. However, one possible approach could be to weight the curation index by proportion of non-flint material in the assemblages. the exlusion of this material The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps. The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel out. Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern, while the high expediency on some sites when considering all lithics in aggregate in the later parts of the Mesolithic does speak to variability in site use. The nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential-logistic continuum. A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014]. However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could in this perspective represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
